---
title: "CV"
---

::: {.content-visible when-profile="english"}
##  🎓 Education

* 2023 — current: PhD in Psychiatry and Behavioral Sciences, Department of
Psychiatry, Federal University of Rio Grande do Sul, Porto Alegre, Rio Grande
do Sul, Brazil
  - Funded by *Coordenação de Aperfeiçoamento de Pessoal de Nível Superior*
  (CAPES) under the Academic Excellence Program (Proex).

* 2023 — 2024: Training in Cognitive-Behavioral Therapies, Institute of
Neuroscience and Cognitive Therapies (INTCognitivas), Porto Alegre, Rio Grande
do Sul, Brazil

* 2022 — 2023: MSc in Psychiatry and Behavioral Sciences, Department of
Psychiatry, Federal University of Rio Grande do Sul, Porto Alegre, Rio Grande
do Sul, Brazil
  - Funded by *Coordenação de Aperfeiçoamento de Pessoal de Nível Superior*
  (Coordination of Superior Level Staff Improvement) under the Academic
  Excellence Program (Proex). Dissertation: [Predição de transtorno bipolar e
  desfechos funcionais em adultos jovens: um acompanhamento de cinco
  anos](http://hdl.handle.net/10183/266476) (English translation: Predicting
  bipolar disorder and functional outcomes in young adults: a five-year
  follow-up). Dissertation defense: [Virtual session
  recording](https://youtu.be/s8ccoTtqgLI)

* 2016 — 2021: BSc in Psychology, Department of Psychology, Catholic University
of Pelotas, Pelotas, Rio Grande do Sul, Brazil
  - Funded by *Conselho Nacional de Desenvolvimento Científico e Tecnológico*
  (CNPq) with Scientific Initiation scholarship. [Final
  paper](https://doi.org/10.1016/j.psycom.2023.100121) was published as an open
  access article at Psychiatry Research Communications.

* 2013 — 2015: Technical Course in Computer Science, Federal Institute of
Education, Science and Technology of Rio Grande do Sul, Ibirubá, Rio Grande do
Sul, Brazil

## 📝 Publications and preprints

* [Clinical and lifestyle predictors of loneliness: A two-year longitudinal
study](https://doi.org/10.1016/j.jpsychires.2024.11.025). Antonelli-Salgado T,
**Montezano BB**, Roza TH, Bouvier V, Zimerman A, Noronha LT, Marcon G,
Hoffmann MS, Brunoni AR, Passos IC. *Article* published at the *Journal of
Psychiatric Research*. 2024.

* [Predictors of response after single session interventions for emotional
distress: using enhanced psychoeducation in crisis
situations](http://dx.doi.org/10.47626/1516-4446-2024-3749). Ache ALS, Baldez
DP, **Montezano BB**, Mosqueiro BP, Caldieraro MA, Spanemberg L, Salum GA,
Fleck MP. *Article* published at the *Brazilian Journal of Psychiatry*. 2024.

* [Subcutaneous ketamine reduces suicide risk and improves functioning in
depression: a proof-of-concept
study](https://doi.org/10.1016/j.psychres.2024.115915). Anzolin AP, Baldez DP,
**Montezano BB**, Kapczinski F, Abreu PB, Kauer-Santanna M. *Article* published
at the *Psychiatry Research*. 2024.

* [The role of parental stress on emotional and behavioral problems in
offspring: a systematic review with
meta-analysis](https://doi.org/10.1016/j.jped.2024.02.003). Ribas LH,
**Montezano BB**, Nieves M, Kampmann LB, Jansen K. *Article* published at
*Jornal de Pediatria*. 2024.

* [Exploring protective factors in a high-risk subsample: the pivotal role of
paternal support in preventing depression in a cohort of young
adults](https://doi.org/10.47626/2237-6089-2024-0804). Tietbohl-Santos B,
**Montezano BB**, Cardoso TA, Mondin TC, Moreira FP, Souza LDM, Silva RA,
Kapczinski F, Jansen K, Passos IC. *Article* published at *Trends in Psychiatry
and Psychotherapy*. 2024.

* [Protective factors against depression in high-risk children and adolescents:
a systematic review of longitudinal
studies](https://doi.org/10.47626/1516-4446-2023-3363). Tietbohl-Santos B,
Shintani AO, **Montezano BB**, Biazin P, Signori GM, Pulice R, Vecchia GFD,
Bebber JC, Noronha LT, Passos IC. *Article* published at the *Brazilian Journal
of Psychiatry*. 2024.

* [Perinatal factors as risk factors of bipolar disorder onset in young
adulthood: a 22-year birth
cohort](https://doi.org/10.47626/1516-4446-2023-3338). Gnielka V, **Montezano
BB**, Baldez DP, Shintani AO, Rabelo-da-Ponte FD, Menezes AMB, Wehrmeister FC,
Gonçalves H, Kunz M, Kauer-Sant'Anna M, Watts D, Kapczinski F, Passos IC.
*Article* published at the *Brazilian Journal of Psychiatry*. 2024.

* [Early exposure to cannabis and bipolar disorder incidence: Findings from a
22-year birth cohort study in Brazil](https://doi.org/10.1111/acps.13670).
Jorge ACR, **Montezano BB**, Aguiar KR, Noronha LT, Baldez DP, Watts D, Menezes
AMB, Wehrmeister F, Gonçalves H, Kunz M, Kapczinski F, Passos IC. *Article*
published at *Acta Psychiatrica Scandinavica*. 2024. *Note:* Alan Cristian
Rodrigues Jorge, Bruno Braga Montezano, and Kyara Rodrigues de Aguiar
contributed equally to the manuscript and are jointly first authors.

* [Autophagy-based
antidepressants?](https://doi.org/10.47626/1516-4446-2023-3127). Kapczinski F,
**Montezano BB**, Nardi AE, Lledo PM, Katsimpardi L. *Editorial* published at
the *Brazilian Journal of Psychiatry*. 2023.

* [Latent bipolar disorder](https://doi.org/10.1016/S0140-6736(23)00402-6).
Kapczinski F, **Montezano BB**, Cardoso TA. *Correspondence* published at *The
Lancet*. 2023.

* [Validity and Reliability of the Brazilian Version of the Cognitive Reserve
Assessment Scale in Health
(CRASH)](https://www.bjp.org.br/details/2417/en-US/validity-and-reliability-of-the-brazilian-version-of-the-cognitive-reserve-assessment-scale-in-health--crash-).
Bücker J, Amoretti S, Vieta E, Czepielewski L, Scheibe DB, **Montezano BB**,
Goularte JF, Rosa AR. *Brief Report* published at the *Brazilian Journal of
Psychiatry*. 2023.

* [Predicting bipolar disorder incidence in young adults using gradient
boosting: a 5-year follow-up
study](https://doi.org/10.1101/2023.03.31.22282507). **Montezano BB**, Gnielka
V, Shintani AO, Aguiar KR, Roza TH, Cardoso TA, Souza LDM, Moreira FP, Silva
RA, Mondin TC, Jansen K, Passos IC. *Preprint*  at *MedRxiv*. 2023.

* [Associations between general sleep quality and measures of functioning and
cognition in subjects recently diagnosed with bipolar
disorder](https://doi.org/10.1016/j.psycom.2023.100121).  **Montezano BB**,
Cardoso TA, Souza LDM, Moreira FP, Mondin TC, Silva RA, Jansen K. *Article*
published at *Psychiatry Research Communications*. 2023.

* [Validation of the Brazilian version of the Processes of Change Questionnaire
in Weight Management in adults with overweight and obesity in
Brazil](https://doi.org/10.1101/2023.03.20.23287450). Carvalho Q, Ziani PR,
**Montezano BB**, Goularte JF, Rosa AR. *Preprint* at *MedRxiv*. 2023.

* [Predictors of illicit substance abuse/dependence during young adulthood: A
machine learning approach](https://doi.org/10.1016/j.jpsychires.2022.11.030).
Rakovski C, Ballester PL, **Montezano BB**, Souza LDM, Jansen K, Silva RA,
Mondin TC, Moreira FP, De Boni RB, Frey BN, Kapczinski F, Cardoso TA. *Article*
published in *Journal of Psychiatric Research*. 2023.

* [Predicting 3-year persistent or recurrent major depressive episode using
machine learning techniques](https://doi.org/10.1016/j.psycom.2022.100055).
Fialho AR, **Montezano BB**, Ballester PL, Cardoso TA, Mondin TC, Moreira FP,
Souza LDM, Silva RA, Jansen K. *Article* published in *Psychiatry Research
Communications*. 2022.

* [Personality disorders as predictors for the conversion from major depressive
disorder to bipolar disorder: A prospective cohort
study](https://doi.org/10.1016/j.psycom.2022.100068). Ávila GB, **Montezano
BB**, Souza LDM, Cardoso TA, Silva RA, Mondin TC, Moreira FP, Jansen K.
*Article* published in *Psychiatry Research Communications*. 2022.

* [Sleep alterations as a predictor of bipolar disorder among offspring of
parents with bipolar disorder: a systematic review and
meta-analysis](https://doi.org/10.47626/2237-6089-2021-0256). Aguiar KR,
Cabelleira MD, **Montezano BB**, Jansen K, Cardoso TA. *Article* published in
*Trends in Psychiatry and Psychotherapy*. 2022.

* [Predicting functional impairment in euthymic patients with mood disorder: A
5-year follow-up](https://doi.org/10.1016/j.psychres.2023.115404). Aguiar KR,
**Montezano BB**, Feiten JG, Watts D, Zimerman A, Mondin TC, Silva RA, Souza
LDM, Kapczinski F, Cardoso TA, Jansen K, Passos, IC. *Article* published in
*Psychiatry Research*. 2023.

## 📚 Book chapters

* [Circadian Rhythms in Patients with Bipolar
Disorder](https://doi.org/10.1007/978-3-031-85519-1_13). **Montezano BB**.
Chapter of the book *Bipolar Disorder: An Evidence-Based Clinical Guide*
published by *Springer Nature*. 2025.

* [Substance Use Disorder in the Course of Bipolar
Disorder](https://doi.org/10.1007/978-3-031-85519-1_11). **Montezano BB**.
Chapter of the book *Bipolar Disorder: An Evidence-Based Clinical Guide*
published by *Springer Nature*. 2025.

* [Prodromal Symptoms and Risk Markers of Bipolar
Disorder](https://doi.org/10.1007/978-3-031-85519-1_5). **Montezano BB**.
Chapter of the book *Bipolar Disorder: An Evidence-Based Clinical Guide*
published by *Springer Nature*. 2025.

* [O uso da inteligência artificial para identificar doenças
psiquiátricas](https://loja.grupoa.com.br/psiquiatria-digital9786558821618-p1021609).
**Montezano BB**, Vecchia GFD, Lorenzoni IC, Passos IC. Chapter of the book
*Psiquiatria Digital* published by *Artmed*. 2024.

* [Predição de risco de suicídio utilizando técnicas de inteligência
artificial](https://loja.grupoa.com.br/psiquiatria-digital9786558821618-p1021609).
Roza TH, **Montezano BB**, Antonelli-Salgado T, Passos IC. Chapter of the book
*Psiquiatria Digital* published by *Artmed*. 2024.

* [The Dawn of Digital
Psychiatry](https://doi.org/10.1007/978-3-031-10698-9_1). Zimerman A,
**Montezano BB**, Vecchia GFD, Kapczinski F, Passos IC. Chapter of the book
*Digital Mental Health* published by *Springer Nature*. 2023.

* [Suicídio](https://loja.grupoa.com.br/psicofarmacos-consulta-rapida-6ed9786558821366-p1021397).
Shintani AO, Salgado TA, Aguiar, KR, Roza TH, Lorenzoni IC, **Montezano BB**,
Passos IC. Chapter of the book *Psicofármacos: consulta rápida* published by
*Artmed*. 2023.

* [Relação entre abuso e dependência de substância e transtorno bipolar em uma
amostra comunitária de jovens](http://dx.doi.org/10.22533/at.ed.400212903).
**Montezano BB**, Jansen K, Cardoso TA, Souza LDM, Silva RA, Mondin TC. Chapter
of the book *Psicologia: Identidade Profissional e Compromisso Social*
published by *Atena Editora*. 2021.

## 🧑‍🏫 Courses 

* [Introduction to R Course (in
Portuguese)](https://brunomontezano.github.io/r-workshop/). Online. April,
2023.
- I conducted a comprehensive Introduction to R course for researchers in the
healthcare field, consisting of seven lectures under the umbrella of the
[AlliaNCE (Artificial Intelligence in Neuroscience) research
group](https://allianceresearchgroup.com/) at UFRGS. The course covered
essential concepts related to data manipulation and analysis using the
open-source programming language R. The curriculum included lectures, practical
exercises, and weekly labs for hands-on application of learned concepts. The
course website provided additional resources, such as recommended materials,
FAQs, lecture slides, codes, and class recordings. Homework assignments with
corresponding answer keys were available, and a tips section featured short
videos illustrating data manipulation and analysis routines. The seven classes
spanned topics from R introduction to machine learning applications, equipping
participants with valuable skills for data-driven research in healthcare. 

## 🎤 Lectures

* [Large Language Models in Mental
Health](https://youtu.be/dnOxqYQJfSE?si=2v0tf3fimgPAvPbX). Porto Alegre, RS.
August 31, 2024.
  - The lecture was given on August 31, 2024 at the Digital Meantal Health
  Meeting, which took place at the Instituto Caldeira in Porto Alegre. The
  lecture covered topics related to new language AI models and how they can be
  used in for mental health related scenarios.

* [Big Data in Mental Health (in
Portuguese)](https://youtu.be/Hduz2YOusiA?si=BZcV0HenNq-kKu8C). Porto Alegre,
RS. August 31, 2024.
  - The lecture was given on August 31, 2024 at the Digital Mental Health
  Meeting, which took place at the Instituto Caldeira in Porto Alegre. The
  lecture covered topics related to the interface between big data and mental
  health, focusing on data science in psychiatry.

* [RISE Project: Stratification of suicide risk in the Brazilian population (in
Portuguese)](https://youtu.be/kWLyciMi7s0?si=cTJ-Ckf6Tya45Xkh). Online. July
29, 2024.
  - The lecture was given on July 29, 2024 via web conference. The lecture
  covered topics related to a multicenter research project that I'm part of.
  The central theme of the seminar was "Suicidal behavior, bipolar disorder and
  borderline personality".

* [Epidemiology and Phenomenology of Bipolar Disorder (in
Portuguese)](https://youtu.be/l_2col7BlNI). Criciúma, SC. September 15, 2023.
  - The lecture was given on September 15, 2023 at 1:30 p.m. at the XXI
  Symposium on Psychiatry at the Brain and Mind Interface, which took place at
  the Universidade do Extremo Sul Catarinense (UNESC) in Criciúma (SC). The
  lecture covered topics related to the epidemiology and phenomenology of
  bipolar disorder. The central theme of the event was "Bipolar disorder: from
  biological bases to treatment".

* [What is Artificial Intelligence? (in
Portuguese)](https://www.instagram.com/p/CwBTdtOpFuB/). Porto Alegre, RS.
August 23, 2023.
  - The lecture was given on August 23, 2023 at an event entitled "Open Class:
  Artificial Intelligence and Digital Psychiatry" organized by the League of
  Psychiatry and Mental Health (LIPSAM) of the Federal University of Health
  Sciences of Porto Alegre (UFCSPA) and the Federal University of Rio Grande do
  Sul (UFRGS). The event took place at TECNOPUC Health Plus in Porto Alegre.
  The lecture covered topics such as big data, machine learning, deep learning,
  the beginnings of AI, as well as aspects of data science applied to
  psychiatry from a historical perspective. It also addressed current
  applications of AI in the field of mental health and how a digital clinic can
  be envisioned for the future.

## 💻 Softwares

* [Dashboard on suicide rates in Brazil from 2000 to
2021](https://brunomontezano.shinyapps.io/dashboard-suicidio/). Written in
Quarto and R with Shiny as backend technology. 2024.
    - Built based on pre-processed data by the [Plataforma de Ciência de Dados
    Aplicada à Saúde](https://pcdas.icict.fiocruz.br/), affiliated to the
    Oswaldo Cruz Foundation.

* [Simple randomization tool for clinical
trials](https://brunomontezano.shinyapps.io/randomizacao-shiny/). Written in
Shiny and R. 2024.
    - This tool provides a solution for randomizing patients in clinical trials
    while maintaining blinding. Additionally, it offers the option for users to
    download tables locally.

## 🏆 Awards

* *[Prêmio Tese FAMED
2024](https://www.ufrgs.br/famed/premio-tese-dissertacao-famed-2024/) (Best
Master's dissertation defended between 01/01/2023 and 12/31/2023, in the
Postgraduate programs of the School of Medicine of the Federal University of
Rio Grande do Sul)*. Federal University of Rio Grande do Sul. Porto Alegre, RS.
2024.

* *Best Poster Award at the [ISBD 2023 Annual
Conference](https://isbd2023.org/)*. International Society for Bipolar
Disorders. Chicago, IL. 2023.

* *Mitacs Globalink Research Award*. Project: Machine learning and women's
mental health: using bleeding-edge methods to understand premenstrual and mood
symptoms. Collaboration with McMaster University (Hamilton, ON).
[Mitacs](https://www.mitacs.ca/en). 2023.

* *Young Researcher Award*, [Congress on Brain, Behavior and Emotions
2022](https://www.braincongress.com.br/). Gramado, RS. 2022.

* *Third place in Best ePoster of the XIII Brazilian Congress of Cognitive
Therapies*, [Brazilian Federation of Cognitive
Therapies](https://fbtc.org.br/). Online. 2021.

* *Highlight in the area of Human Sciences at the 28th Congress of Scientific
Initiation*, [Federal University of Rio Grande](https://www.furg.br/en/). Rio
Grande, RS. 2019.

## 📰 Media and press releases

* [Solidão: prevalência global acende alerta para impactos na saúde da
população e busca por
soluções](https://futurodasaude.com.br/solidao-solucoes/). Futuro da Saúde. May
5, 2025.

* [Prêmio Tese FAMED 2024: Conheça os
vencedores](https://www.ufrgs.br/famed/premio-tese-dissertacao-famed-2024/).
Universidade Federal do Rio Grande do Sul. December 11, 2024.

* [Egressos do curso de Psicologia da UCPel são destaque em premiação
internacional.](https://ucpel.edu.br/noticias/egressos-do-curso-de-psicologia-da-ucpel-sao-destaque-em-premiacao-internacional)
Universidade Católica de Pelotas. June 29, 2023.

* [Estudo de pesquisadores da UFRGS e da França identifica uma possível causa
neurobiológica da
depressão.](https://gauchazh.clicrbs.com.br/saude/vida/noticia/2023/04/estudo-de-pesquisadores-da-ufrgs-e-da-franca-identifica-uma-possivel-causa-neurobiologica-da-depressao-clgf3u4kn000h015xcdaveosq.html)
Zero Hora. April 13, 2023.

## 🧑‍💼 Roles

* Assistant Editor: [Brazilian Journal of Psychiatry](https://www.bjp.org.br/)
* Reviewer:
    - [Brazilian Journal of Psychiatry](https://www.bjp.org.br/)
    - [Nature Scientific Reports](https://www.nature.com/srep/)
    - [Journal of Medical Internet Research](https://www.jmir.org/)
    - [The R Journal](https://journal.r-project.org/)
    - [The Journal of Open Source Software](https://joss.theoj.org/)
    - [Debates em Psiquiatria](https://revistardp.org.br/)
    - [Trends in Psychiatry and Psychotherapy](https://www.scielo.br/j/trends/)
    - [International Journal of Mental Health and Addiction](https://www.springer.com/journal/11469)

## 📈 Citations

```{r citation-plotly-en-us}
#| echo: false
citations <- scholar::get_citation_history("dULhWr0AAAAJ") |>
  dplyr::mutate(text = glue::glue("Year: {year}\nCitations: {cites}"))

citations_plot <- plotly::plot_ly(
  data = citations,
  x = ~year,
  y = ~cites,
  text = ~text,
  hoverinfo = "text",
  textposition = "none",
  name = "Google Scholar cites",
  type = "bar",
  marker = list(
    color = "rgb(73, 150, 245)",
    line = list(color = "rgb(0, 0, 0)", width = 1.5)
  )
) |>
  plotly::layout(
    yaxis = list(
      title = "Google Scholar cites",
      titlefont = list(size = 15),
      fixedrange = TRUE
    ),
    xaxis = list(
      title = "Year",
      titlefont = list(size = 15),
      tickvals = as.list(2022:2024),
      tickmode = "array",
      fixedrange = TRUE
    )
  ) |>
  plotly::add_annotations(
    x = min(citations$year),
    y = max(citations$cites) + 1,
    text = paste0("Last updated: ", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")),
    showarrow = FALSE,
    font = list(color = "grey90")
  ) |>
  plotly::config(
    displayModeBar = TRUE,
    displaylogo = FALSE,
    modeBarButtonsToRemove = list(
      "lasso2d",
      "select2d",
      "hoverClosestCartesian",
      "hoverCompareCartesian"
    )
  )

citations_plot
```
:::


::: {.content-visible when-profile="portuguese"}
##  🎓 Formação acadêmica

* 2023 — atual: Doutorado em Psiquiatria e Ciências do Comportamento,
Departamento de Psiquiatria, Universidade Federal do Rio Grande do Sul, Porto
Alegre, Rio Grande do Sul, Brasil
  - Financiado pela *Coordenação de Aperfeiçoamento de Pessoal de Nível Superior*
  (CAPES) através do Programa de Excelência Acadêmica (Proex).

* 2023 — 2024: Formação em Terapias Cognitivo-Comportamentais, Instituto de
Neurociências e Terapias Cognitivas (INTCognitivas), Porto Alegre, Rio Grande
do Sul, Brasil

* 2022 — 2023: Mestrado em Psiquiatria e Ciências do Comportamento, Departamento de
Psiquiatria, Universidade Federal do Rio Grande do Sul, Porto Alegre, Rio Grande
do Sul, Brasil
  - Financiado pela *Coordenação de Aperfeiçoamento de Pessoal de Nível Superior*
  (CAPES) através do Programa de Excelência Acadêmica
  (Proex). Dissertação: [Predição de transtorno bipolar e
  desfechos funcionais em adultos jovens: um acompanhamento de cinco
  anos](http://hdl.handle.net/10183/266476) (Em inglês: Predicting
  bipolar disorder and functional outcomes in young adults: a five-year
  follow-up). Defesa de dissertação: [Acesso via YouTube](https://youtu.be/s8ccoTtqgLI).

* 2016 — 2021: Graduação em Psicologia, Departamento de Psicologia, Universidade Católica de Pelotas, Pelotas, Rio Grande do Sul, Brasil
  - Financiado pelo *Conselho Nacional de Desenvolvimento Científico e Tecnológico*
  (CNPq) com Bolsa de Iniciação Científica. [Trabalho de  conclusão  de curso](https://doi.org/10.1016/j.psycom.2023.100121) foi publicado
  como artigo de acesso livre na revista *Psychiatry Research Communications*.

* 2013 — 2015: Curso Técnico em Informática, Instituto Federal de Educação,
Ciência e Tecnologia do Rio Grande do Sul, Ibirubá, Rio Grande do Sul, Brasil

## 📝 Publicações

* [Clinical and lifestyle predictors of loneliness: A two-year longitudinal
study](https://doi.org/10.1016/j.jpsychires.2024.11.025). Antonelli-Salgado T,
**Montezano BB**, Roza TH, Bouvier V, Zimerman A, Noronha LT, Marcon G,
Hoffmann MS, Brunoni AR, Passos IC. *Artigo* publicado no *Journal of
Psychiatric Research*. 2024.

* [Predictors of response after single session interventions for emotional
distress: using enhanced psychoeducation in crisis
situations](http://dx.doi.org/10.47626/1516-4446-2024-3749). Ache ALS, Baldez
DP, **Montezano BB**, Mosqueiro BP, Caldieraro MA, Spanemberg L, Salum GA,
Fleck MP. *Artigo* publicado no *Brazilian Journal of Psychiatry*. 2024.

* [Subcutaneous ketamine reduces suicide risk and improves functioning in
depression: a proof-of-concept
study](https://doi.org/10.1016/j.psychres.2024.115915). Anzolin AP, Baldez DP,
**Montezano BB**, Kapczinski F, Abreu PB, Kauer-Santanna M. *Artigo* publicado
na revista *Psychiatry Research*. 2024.

* [The role of parental stress on emotional and behavioral problems in
offspring: a systematic review with
meta-analysis](https://doi.org/10.1016/j.jped.2024.02.003). Ribas LH,
**Montezano BB**, Nieves M, Kampmann LB, Jansen K. *Artigo* publicado no
*Jornal de Pediatria*. 2024.

* [Exploring protective factors in a high-risk subsample: the pivotal role of
paternal support in preventing depression in a cohort of young
adults](https://doi.org/10.47626/2237-6089-2024-0804). Tietbohl-Santos B,
**Montezano BB**, Cardoso TA, Mondin TC, Moreira FP, Souza LDM, Silva RA,
Kapczinski F, Jansen K, Passos IC. *Artigo* publicado na *Trends in Psychiatry
and Psychotherapy*. 2024.

* [Protective factors against depression in high-risk children and adolescents:
a systematic review of longitudinal
studies](https://doi.org/10.47626/1516-4446-2023-3363). Tietbohl-Santos B,
Shintani AO, **Montezano BB**, Biazin P, Signori GM, Pulice R, Vecchia GFD,
Bebber JC, Noronha LT, Passos IC. *Artigo* publicado no *Brazilian Journal
of Psychiatry*. 2024.

* [Perinatal factors as risk factors of bipolar disorder onset in young
adulthood: a 22-year birth
cohort](https://doi.org/10.47626/1516-4446-2023-3338). Gnielka V, **Montezano
BB**, Baldez DP, Shintani AO, Rabelo-da-Ponte FD, Menezes AMB, Wehrmeister FC,
Gonçalves H, Kunz M, Kauer-Sant'Anna M, Watts D, Kapczinski F, Passos IC.
*Artigo* publicado no *Brazilian Journal of Psychiatry*. 2024.

* [Early exposure to cannabis and bipolar disorder incidence: Findings from a
22-year birth cohort study in Brazil](https://doi.org/10.1111/acps.13670).
Jorge ACR, **Montezano BB**, Aguiar KR, Noronha LT, Baldez DP, Watts D, Menezes
AMB, Wehrmeister F, Gonçalves H, Kunz M, Kapczinski F, Passos IC. *Artigo*
publicado na revista *Acta Psychiatrica Scandinavica*. 2024. *Nota:* Alan
Cristian
Rodrigues Jorge, Bruno Braga Montezano e Kyara Rodrigues de Aguiar contribuíam
igualmente ao manuscrito e são primeiros autores em conjunto.

* [Autophagy-based
antidepressants?](https://doi.org/10.47626/1516-4446-2023-3127). Kapczinski F,
**Montezano BB**, Nardi AE, Lledo PM, Katsimpardi L. *Editorial* publicado no
*Brazilian Journal of Psychiatry*. 2023.

* [Latent bipolar disorder](https://doi.org/10.1016/S0140-6736(23)00402-6).
Kapczinski F, **Montezano BB**, Cardoso TA. *Correspondence* publicada no *The
Lancet*. 2023.

* [Validity and Reliability of the Brazilian Version of the Cognitive Reserve
Assessment Scale in Health
(CRASH)](https://www.bjp.org.br/details/2417/en-US/validity-and-reliability-of-the-brazilian-version-of-the-cognitive-reserve-assessment-scale-in-health--crash-).
Bücker J, Amoretti S, Vieta E, Czepielewski L, Scheibe DB, **Montezano BB**,
Goularte JF, Rosa AR. *Brief Report* publicado no *Brazilian Journal of
Psychiatry*. 2023.

* [Predicting bipolar disorder incidence in young adults using gradient
boosting: a 5-year follow-up
study](https://doi.org/10.1101/2023.03.31.22282507). **Montezano BB**, Gnielka
V, Shintani AO, Aguiar KR, Roza TH, Cardoso TA, Souza LDM, Moreira FP, Silva
RA, Mondin TC, Jansen K, Passos IC. *Preprint* disponível no *MedRxiv*. 2023.

* [Associations between general sleep quality and measures of functioning and
cognition in subjects recently diagnosed with bipolar
disorder](https://doi.org/10.1016/j.psycom.2023.100121).  **Montezano BB**,
Cardoso TA, Souza LDM, Moreira FP, Mondin TC, Silva RA, Jansen K. *Artigo*
publicado na revista *Psychiatry Research Communications*. 2023.

* [Validation of the Brazilian version of the Processes of Change Questionnaire
in Weight Management in adults with overweight and obesity in
Brazil](https://doi.org/10.1101/2023.03.20.23287450). Carvalho Q, Ziani PR,
**Montezano BB**, Goularte JF, Rosa AR. *Preprint* disponível no *MedRxiv*. 2023.

* [Predictors of illicit substance abuse/dependence during young adulthood: A
machine learning approach](https://doi.org/10.1016/j.jpsychires.2022.11.030).
Rakovski C, Ballester PL, **Montezano BB**, Souza LDM, Jansen K, Silva RA,
Mondin TC, Moreira FP, De Boni RB, Frey BN, Kapczinski F, Cardoso TA. *Artigo*
publicado no *Journal of Psychiatric Research*. 2023.

* [Predicting 3-year persistent or recurrent major depressive episode using
machine learning techniques](https://doi.org/10.1016/j.psycom.2022.100055).
Fialho AR, **Montezano BB**, Ballester PL, Cardoso TA, Mondin TC, Moreira FP,
Souza LDM, Silva RA, Jansen K. *Artigo* publicado na revista *Psychiatry
Research Communications*. 2022.

* [Personality disorders as predictors for the conversion from major depressive
disorder to bipolar disorder: A prospective cohort
study](https://doi.org/10.1016/j.psycom.2022.100068). Ávila GB, **Montezano
BB**, Souza LDM, Cardoso TA, Silva RA, Mondin TC, Moreira FP, Jansen K.
*Artigo* publicado na revista *Psychiatry Research Communications*. 2022.

* [Sleep alterations as a predictor of bipolar disorder among offspring of
parents with bipolar disorder: a systematic review and
meta-analysis](https://doi.org/10.47626/2237-6089-2021-0256). Aguiar KR,
Cabelleira MD, **Montezano BB**, Jansen K, Cardoso TA. *Artigo* publicado na
*Trends in Psychiatry and Psychotherapy*. 2022.

* [Predicting functional impairment in euthymic patients with mood disorder: A
5-year follow-up](https://doi.org/10.1016/j.psychres.2023.115404). Aguiar KR,
**Montezano BB**, Feiten JG, Watts D, Zimerman A, Mondin TC, Silva RA, Souza
LDM, Kapczinski F, Cardoso TA, Jansen K, Passos, IC. *Artigo* publicado na
revista *Psychiatry Research*. 2023.

## 📚 Capítulos de livro

* [Circadian Rhythms in Patients with Bipolar
Disorder](https://doi.org/10.1007/978-3-031-85519-1_13). **Montezano BB**.
Capítulo do livro *Bipolar Disorder: An Evidence-Based Clinical Guide*
publicado pela *Springer Nature*. 2025.

* [Substance Use Disorder in the Course of Bipolar
Disorder](https://doi.org/10.1007/978-3-031-85519-1_11). **Montezano BB**.
Capítulo do livro *Bipolar Disorder: An Evidence-Based Clinical Guide*
publicado pela *Springer Nature*. 2025.

* [Prodromal Symptoms and Risk Markers of Bipolar
Disorder](https://doi.org/10.1007/978-3-031-85519-1_5). **Montezano BB**.
Capítulo do livro *Bipolar Disorder: An Evidence-Based Clinical Guide*
publicado pela *Springer Nature*. 2025.

* [O uso da inteligência artificial para identificar doenças
psiquiátricas](https://loja.grupoa.com.br/psiquiatria-digital9786558821618-p1021609).
**Montezano BB**, Vecchia GFD, Lorenzoni IC, Passos IC. Capítulo do livro
*Psiquiatria Digital* publicado pela *Artmed*. 2024.

* [Predição de risco de suicídio utilizando técnicas de inteligência
artificial](https://loja.grupoa.com.br/psiquiatria-digital9786558821618-p1021609).
Roza TH, **Montezano BB**, Antonelli-Salgado T, Passos IC. Capítulo do livro
*Psiquiatria Digital* publicado pela *Artmed*. 2024.

* [The Dawn of Digital
Psychiatry](https://doi.org/10.1007/978-3-031-10698-9_1). Zimerman A,
**Montezano BB**, Vecchia GFD, Kapczinski F, Passos IC. Capítulo do livro
*Digital Mental Health* publicado pela *Springer Nature*. 2023.

* [Suicídio](https://loja.grupoa.com.br/psicofarmacos-consulta-rapida-6ed9786558821366-p1021397).
Shintani AO, Salgado TA, Aguiar, KR, Roza TH, Lorenzoni IC, **Montezano BB**,
Passos IC. Capítulo do livro *Psicofármacos: consulta rápida* publicado pela
*Artmed*. 2023.

* [Relação entre abuso e dependência de substância e transtorno bipolar em uma
amostra comunitária de jovens](http://dx.doi.org/10.22533/at.ed.400212903).
**Montezano BB**, Jansen K, Cardoso TA, Souza LDM, Silva RA, Mondin TC.
Capítulo do livo *Psicologia: Identidade Profissional e Compromisso Social*
publicado pela *Atena Editora*. 2021.

## 🧑‍🏫 Cursos gratuitos 

* [Introdução ao R: aplicações em ciências da
saúde](https://brunomontezano.github.io/r-workshop/). Online. Abril de 2023.
    - Conduzi um curso introdutório sobre a linguagem de programação R para
    pesquisadores da área da saúde, composto por sete aulas, para os
    integrantes do [grupo de pesquisa AlliaNCE (Artificial Intelligence in
    Neuroscience)](https://allianceresearchgroup.com/) da UFRGS. O curso
    abordou conceitos essenciais relacionados à manipulação e
    análise de dados utilizando a linguagem de programação R. O currículo
    incluiu aulas teóricas, exercícios práticos e laboratórios semanais para
    aplicação prática dos conceitos aprendidos. O site do curso fornece
    recursos adicionais, como materiais recomendados, perguntas frequentes,
    slides das aulas, códigos e gravações das aulas. Foram disponibilizadas
    listas de exercícios com gabaritos correspondentes, além de uma seção de
    dicas com vídeos curtos ilustrando rotinas de manipulação e análise de
    dados. As sete aulas abordaram temas desde a introdução ao R até aplicações
    de aprendizado de máquina, proporcionando aos participantes habilidades
    valiosas para pesquisa orientada por dados na área da saúde.

## 🎤 Palestras

* [Large Language Models na Saúde
Mental](https://youtu.be/dnOxqYQJfSE?si=2v0tf3fimgPAvPbX). Porto Alegre, RS. 31
de agosto de 2024.
    - A palestra foi realizada em 31 de agosto de 2024, durante o Meeting de
    Saúde Mental Digital, que ocorreu no Instituto Caldeira, em Porto Alegre. A
    apresentação abordou temas relacionados aos novos modelos de linguagem
    baseados em inteligência artificial e como eles podem ser utilizados em
    cenários relacionados à saúde mental.

* [Big Data na Saúde Mental](https://youtu.be/Hduz2YOusiA?si=BZcV0HenNq-kKu8C). Porto Alegre,
RS. 31 de agosto de 2024.
    - A palestra foi realizada em 31 de agosto de 2024, durante o Meeting de
    Saúde Mental Digital, que ocorreu no Instituto Caldeira, em Porto Alegre. A
    apresentação abordou temas relacionados à interface entre big data e saúde
    mental, com foco na ciência de dados em psiquiatria.

* [Projeto RISE: Estratificação do risco de suicídio na população
brasileira](https://youtu.be/kWLyciMi7s0?si=cTJ-Ckf6Tya45Xkh). Online. 29 de
julho de 2024.
    - A palestra foi realizada em 29 de julho de 2024, por meio de
    webconferência. A apresentação abordou temas relacionados a um projeto de
    pesquisa multicêntrico do qual faço parte. O tema central do seminário foi
    "Comportamento suicida, transtorno bipolar e transtorno de personalidade
    borderline". O evento foi organizado por pesquisadores da Universidade
    Federal de Sergipe.

* [Epidemiologia e Fenomenologia do Transtorno
Bipolar](https://youtu.be/l_2col7BlNI). Criciúma, SC. 15 de setembro de 2023.
    - A palestra foi realizada em 15 de setembro de 2023, às 13h30, durante o
    XXI Simpósio de Psiquiatria na Interface Cérebro e Mente, que ocorreu na
    Universidade do Extremo Sul Catarinense (UNESC), em Criciúma (SC). A
    apresentação abordou temas relacionados à epidemiologia e fenomenologia do
    transtorno bipolar. O tema central do evento foi "Transtorno bipolar: das
    bases biológicas ao tratamento".

* [O que é Inteligência Artificial?](https://www.instagram.com/p/CwBTdtOpFuB/).
Porto Alegre, RS. 23 de agosto de 2023.
    - A palestra foi realizada em 23 de agosto de 2023, durante o evento
    intitulado "Aula Aberta: Inteligência Artificial e Psiquiatria Digital",
    organizado pela Liga de Psiquiatria e Saúde Mental (LIPSAM) da Universidade
    Federal de Ciências da Saúde de Porto Alegre (UFCSPA) e da Universidade
    Federal do Rio Grande do
    Sul (UFRGS). O evento ocorreu no TECNOPUC Health Plus, em Porto Alegre. A
    apresentação abordou temas como big data, machine learning, deep learning,
    os primórdios da IA, bem como aspectos da ciência de dados aplicada à
    psiquiatria sob uma perspectiva histórica. Também foram discutidas as
    aplicações atuais da IA na área da saúde mental e como pode ser concebida
    uma clínica digital no futuro.

## 💻 Softwares

* [Dashboard de taxas de suicídio no Brasil: de 2000 a
2021](https://brunomontezano.shinyapps.io/dashboard-suicidio/). Escrito em
Quarto e R via Shiny. 2024.
    - Construído a partir de dados pré-processados pela [Plataforma de Ciência
    de Dados Aplicada à Saúde](https://pcdas.icict.fiocruz.br/), vinculada a
    Fundação Oswaldo Cruz (FIOCRUZ).
    - *Nota*: atualmente indisponível por questões de sigilo de dados.

* [Ferramenta de randomização para ensaios
clínicos](https://brunomontezano.shinyapps.io/randomizacao-shiny/). Escrito em
Shiny e R. 2024.
    - A ferramenta soluciona a randomização simples de pacientes em ensaios
    clínicos, mantendo o processo de cegamento. Ainda, oferece aos usuários a
    opção de baixar as tabelas localmente.

## 🏆 Prêmios e destaques

* *[Prêmio Tese FAMED
2024](https://www.ufrgs.br/famed/premio-tese-dissertacao-famed-2024/) (Melhor
dissertação defendida entre 2023 e 2024 nos programas de pós-graduação da
Faculdade de Medicina da UFRGS)*. Universidade Federal do Rio Grande do Sul.
Porto Alegre, RS. 2024.

* *Best Poster Award no [ISBD 2023 Annual
Conference](https://isbd2023.org/)*. International Society for Bipolar
Disorders. Chicago, IL. 2023.

* *Mitacs Globalink Research Award*. Projeto: Machine learning and women's
mental health: using bleeding-edge methods to understand premenstrual and mood
symptoms. Colaboração com McMaster University (Hamilton, ON, Canadá).
[Mitacs](https://www.mitacs.ca/en). 2023.

* *Prêmio Jovem Pesquisador*, [Congress on Brain, Behavior and Emotions
2022](https://www.braincongress.com.br/). Gramado, RS. 2022.

* *Terceiro lugar de Melhor ePôster do XIII Congresso Brasileiro de Terapias
Cognitivas*, [Federação Brasileira de Terapias
Cognitivas](https://fbtc.org.br/). Online. 2021.

* *Destaque na área de Ciências Humanas no 28 Congresso de
Iniciação Científica*, [Universidade Federal de Rio
Grande](https://www.furg.br/en/). Rio Grande, RS. 2019.

## 📰 Entrevistas e comunicados de imprensa

* [Solidão: prevalência global acende alerta para impactos na saúde da
população e busca por
soluções](https://futurodasaude.com.br/solidao-solucoes/). Futuro da Saúde. 5 de maio de 2025.

* [Prêmio Tese FAMED 2024: Conheça os
vencedores](https://www.ufrgs.br/famed/premio-tese-dissertacao-famed-2024/).
Universidade Federal do Rio Grande do Sul. 11 de dezembro de 2024.

* [Egressos do curso de Psicologia da UCPel são destaque em premiação
internacional.](https://ucpel.edu.br/noticias/egressos-do-curso-de-psicologia-da-ucpel-sao-destaque-em-premiacao-internacional)
Universidade Católica de Pelotas. 29 de junho de 2023.

* [Estudo de pesquisadores da UFRGS e da França identifica uma possível causa
neurobiológica da
depressão.](https://gauchazh.clicrbs.com.br/saude/vida/noticia/2023/04/estudo-de-pesquisadores-da-ufrgs-e-da-franca-identifica-uma-possivel-causa-neurobiologica-da-depressao-clgf3u4kn000h015xcdaveosq.html)
Zero Hora. 13 de abril de 2023.

## 🧑‍💼 Cargos editoriais

* Editor Assistente: [Brazilian Journal of Psychiatry](https://www.bjp.org.br/)
* Revisor:
    - [Brazilian Journal of Psychiatry](https://www.bjp.org.br/)
    - [Nature Scientific Reports](https://www.nature.com/srep/)
    - [Journal of Medical Internet Research](https://www.jmir.org/)
    - [The R Journal](https://journal.r-project.org/)
    - [The Journal of Open Source Software](https://joss.theoj.org/)
    - [Debates em Psiquiatria](https://revistardp.org.br/)
    - [Trends in Psychiatry and Psychotherapy](https://www.scielo.br/j/trends/)
    - [International Journal of Mental Health and Addiction](https://www.springer.com/journal/11469)

## 📈 Citações

```{r citation-plotly-pt-br}
#| echo: false
citations <- scholar::get_citation_history("dULhWr0AAAAJ") |>
  dplyr::mutate(text = glue::glue("Ano: {year}\nCitações: {cites}"))

citations_plot <- plotly::plot_ly(
  data = citations,
  x = ~year,
  y = ~cites,
  text = ~text,
  hoverinfo = "text",
  textposition = "none",
  name = "Citações no Google Scholar",
  type = "bar",
  marker = list(
    color = "rgb(73, 150, 245)",
    line = list(color = "rgb(0, 0, 0)", width = 1.5)
  )
) |>
  plotly::layout(
    yaxis = list(
      title = "Citações no Google Scholar",
      titlefont = list(size = 15),
      fixedrange = TRUE
    ),
    xaxis = list(
      title = "Ano",
      titlefont = list(size = 15),
      tickvals = as.list(2022:2024),
      tickmode = "array",
      fixedrange = TRUE
    )
  ) |>
  plotly::add_annotations(
    x = min(citations$year),
    y = max(citations$cites) + 1,
    text = paste0(
      "Atualizado em: ",
      format(Sys.time(), "%d/%m/%Y %H:%M:%S UTC%Z")
    ),
    showarrow = FALSE,
    font = list(color = "grey90")
  ) |>
  plotly::config(
    displayModeBar = TRUE,
    displaylogo = FALSE,
    modeBarButtonsToRemove = list(
      "lasso2d",
      "select2d",
      "hoverClosestCartesian",
      "hoverCompareCartesian"
    )
  )

citations_plot
```
:::
